visualFields 1.x

### 2024-08-17 vignette.Rmd
IMF revision. Did some general editing, added references, a classic Hill of Vision figure, and an example with Octopus G1 grid and made up data

---
title: "VF1 Vignette"
author: "Cord Huchzermeyer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(visualFields)

example1 <- vffilter(vfpwgSunyiu24d2, 
                     id == "sample1" & 
                       eye == "OD" & 
                       date == min(vfpwgSunyiu24d2$date))

# example1_dataframe <- data.frame(x = getlocmap()$coord$x, y = getlocmap()$coord$y, td = as.vector(unlist(example1[1, 11:64])))

The visualFields package facilitates statistical analysis and plotting of visual field data, focussing on static perimetry.

# What is visual field testing?

Visual field testing measures a patient's visual ability at different locations of the visual field. This ability is usually highest in the center of the visual field (i.e. the point that the observer looks at) and decreases toward the periphery (*hill of vision*). The most common type of visual field testing is *perimetry*, which is performed with in a dome-shaped cupule, so that the distance between the stimulus and the eye equal for all locations within the visual field. In contrast, *campimetry* is performed with a flat screen.

Usually, perimetry tests the ability to perceive differences in luminance with an achromatic (not colored) stimulus (white-on-white perimetry). However, theoretically, other aspects of vision can be tested at different locations (ability to discriminate dots that are very close to each other, ability to discriminate similar colors, contrast sensitivity).

@@

plot(x = default_locmap$coord$x, 
     y = default_locmap$coord$y, 
     type = "point", 
     asp = 1,
     xlim = c(-32, 26),
     ylim = c(-29, 26),
     yaxt = 'n',
     xaxt = 'n',
     bty = 'n',
     xlab = "",
     ylab = "")
text(x = 0, y = -28, "horizontal eccentricity [degree]")
text(x = -34, y = -0, "vertical eccentricity [degree]", srt = 90)
points(x = default_locmap$coord[default_locmap$bs, "x"],
       y = default_locmap$coord[default_locmap$bs, "y"],
       col = "red")
lines(x = c(-2.5, 2.5), y = c(0, 0))
lines(x = c(0, 0), y = c(-2.5, 2.5))

@@

The visual fields package was created to analyze data from *static perimetry*, where a threshold is measured for each of a fixed number of locations in a grid pattern. In contrast, in *kinetic perimetry*, a light spot moves from the periphery toward the center along a number of meridians and the location there the spot is first perceivable for the subject is noted.

# Performing visual field tests

For the typical visual field test, the subject or patient is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld cupule (or hemisphere). The perimeter is located in a dimly lit, calm room and the subject has been explained what to do by an experienced technician before the examination. In his hand, the subject has a button.

He now looks at a fixation light right in front of him in the center of the cupule and presses the button as soon as he sees a light appearing anywhere in the cupule. These lights are presented only for a short time span (typically 100-200ms), so that the subject has no time to shift his gaze to the stimulus. A press of the button will be linked to a presentation when it occurs within a given time window after the presentation. A computer will present stimuli several times at different locations in a grid pattern. For each location, the stimulus intensity (luminance) will be adjusted depending on the subjects responses, but the presentations will be randomly interleaved between different locations.

## Grid pattern

The spatial position of the points $l_1$ to $l_n$ in the visual field is found in the location map (locmap) and depends on the pattern used by the perimeter. When analyzing data, you need to specify the locmap.

By default, the 24-2 pattern (shown above) is used, which consists of 54 locations that are equally spaced across the field.

@@

# install.packages("visualFields")
library(visualFields)

str(default_locmap)

Beside a name and a description, a locmap dataset contains a coordinate table with the x- and y- coordinates of the locations measured in degrees, and the id of the points that cover the physiological blind spot (the id is the number x in $l_x$ and corresponds to the row in the coord table).

ten_two_locmap <- locmaps$p10d2

str(ten_two_locmap)

```
The grid, the stimulus size, presentation time and the algorithm responsible for varying intensities remain constant throughout the examination and are called a perimetry program. Not that examinations performed with different programs cannot be directly compared.

# Clinical applications of visual field tests

In clinical practice, visual field testing is used identify decreases in visual function caused by disease. Such *visual field defects* or *scotomas* can be caused by diseases that affect tissue anywhere along the optic pathway. This ranges from the optic media (cornea / lens / vitreous) which transmit light, over the retina which detects lights and preprocesses visual information, the optic nerve which transmits information, up to the brain.

Often, clinicians can draw reliable conclusions about the localisation of the disease by the pattern of the visual field defects. For example, an infarction of the right visual cortex leads complete blindness in the left hemispheres of both eyes with a very distinct border along the vertical meridian. Therefore, the patient does not perceive anything to the right of the point they look at. On the other side of vertical meridian the visual field is normal. In contrast, patients with glaucoma start with areas of slightly perceived light sensitivity often in the upper or lower hemifield with a bow-shaped pattern cause by the localisation of damage in the optic nerve head.

Visual field tests can be used to diagnose disease, to appraise the functional consequences of disease (for example: can the patient still drive a vehicle?), and to monitor progression of disease.

The most advanced statistical methods have been created for monitoring glaucoma progression. The open-angle glaucomas are a group of slowly progressing, chronic neurodegenerative diseases of the optic nerve, often caused by increased eye pressure. Patients need consequent monitoring over many years and slight changes in the visual field must be detected reliably. Therefore, many of the functions of the visualField package focus on glaucoma research.

# Getting started with the package

## Loading data

The visualField package contains some example data. This dataset contains the results of 42 visual field tests of one patient (id: sample1). The patient has primary open-angle glaucoma (type: pwg).

data(vfpwgSunyiu24d2)

@@

Each row corresponds to one visual field test. Visual field testing is performed monocularly (one eye at a time), so that two examinations exist for each date. The abbreviations for the laterality ('eye')is derived from latin (OD: right eye = oculus dexter, and OS: left eye = oculus sinister), when both eyes are examined, the abbreviation is OU.

@@

vfpwgSunyiu24d2[1, getlocini():ncol(vfpwgSunyiu24d2)]

@@

## Plotting the visual field

The most straightforward plots show the sensitivities at the corresponding locations. These representations are very exact, but may be difficult to interpret. For learning purposes, here, we manually plot some visual field data.

@@

example2 <- vffilter(vfpwgSunyiu24d2, 
                     id == "sample1" & 
                       eye == "OD")[1, ]

@@

example2_dataframe <- data.frame(x = getlocmap()$coord$x, 
                                 y = getlocmap()$coord$y, 
                                 td = as.vector(unlist(example2[, getvfcols()])))

ggplot(example2_dataframe, aes(x = x, y = y)) +
  geom_text(aes(label = td)) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  theme_void()

```
Note that left eyes have to be treated differently if you plot the points manually, because the location labels are mirrored.

@@

ggplot(ODOS_dataframe, aes(x = x, y = y)) +
  geom_label(aes(label = loc_label, color = is_blindspot)) +
  scale_color_manual(values = c("black", "red")) +
  facet_wrap(~ eye)

@@

# Psychophysical testing

Psychophysical tests systematically measure the relationship between physical stimulus strength and the subjective perception. This is performed by varying the stimulus and asking the observer to communicate his perception with a limited number of allowed responses. In visual field testing, the observer usually presses a button when a light spot is perceived.

@@

# Measurements

Although usually not seen in visual field data, the most elementary unit of a static visual field test is a subject's response. It consists of the constant stimulus parameters (size, presentation time, background luminance), the stimulus location, the stimulus intensity and whether the subject pressed the button within a predefined time window to indicate that he saw the stimulus.

@@

## False negatives / positives

@@

# Mathematical derivatives of threshold

## Sensitivity and other mathematical derivatives of threshold

However, the data analyzed with the visual fields package usually consists of the subjects *differential light sensitivity* for a stimulus at a given location in decibel (dB) as the most elementary unit. This sensitivity is calculated from the psychophysical threshold, which is derived from the aforementioned responses at this location.

The light sensitivity [dB] is defined as $10*log_{10}{\frac{L_{max}}{L}}$. Where $L$ is the threshold luminance and $L_{max}$ is the maximal stimulus intensity the perimeter uses.

The light sensitivity fluctuates considerably between measurements in the same observer, between observers, but it also decreases with age.

## Defect / Deviation

Clinically it is often more interesting to map the deviation from a theoretical "normal" sensitivity than the absolute sensitivity, because this facilitates the identification of disease. Furthermore, a decline of DLS with age occurs also in normal subjects. Therefore, age-related normative date must be collected which results in a function the describes the relationship between DLS and age for each location in the visual field. This function is usually linear.

The resulting difference is also measured in dB and is called either *deviation* (loss of function has a negative sign; used in Humphrey perimeters) or *defect* (loss of function has a positive sign; used in Octopus perimeters). In this package, *deviation* is used, and it is called *total deviation*. Make sure that you get the right sign if you import data.

In eye diseases like glaucoma, general defects that affect the whole visual field are often less relevant than focal scotomas. Therefore, deviation is sometimes corrected for the average sensitivity of the visual field. This is called *pattern deviation*.

### Sensitivity

In the visualField datasets, $l_1$ to $l_n$ represents sensitivities.

example2[, getlocini()+1:4]

@@

### Total deviation

Total deviation is calculated from the sensitivities and from age-related normal values. These are saved in the environment variable *nv* (normal values).

@@

age <- example2[1, "age"]

@@

normal_values <- getnv()$agem$model(age)[1:4]

@@

total_defects <- example2[, getlocini()+1:4] - normal_values

@@

gettd(example2)[, getlocini()+1:4]

@@

### Pattern deviation

The pattern deviation corrects for the mean deviation, i.e. the average sensitivity loss. It highlights local defects but does not show global sensitivity loss, for example due to opacification of the optic media.

@@

getpd(example2)[, getlocini()+1:4]

@@

## Probabilities

Psychophysical values vary considerably between subjects, but they also fluctuate between measurements. Therefore, deviation will be different from zero most of the time, and the clinician has to know, if a given deviation is likely to be caused by disease, or whether it may just reflect normal variability.

This is calculated based on measurements in the reference population that is also used to calculate the normative data.

### Total deviation

```{r total deviation probabilities}

gettdp(example2)[, getlocini()+1:4]

@@

### Pattern deviation

```{r pattern deviation probabilities}

getpdp(vfpwgSunyiu24d2)[1:4, 1:12]

# Plotting the visual field

There are different ways to visualize perimetric results. These visualizations have to be interpreted like maps in that they represent a simplified version of the hill of vision.

The standard already contains some additional information with a border presenting a color-coded visualization of the deviation and a graphical representation of the blind spot. Note that the vfplot with the option "td" needs information on differential light-sensitivities in a normal reference population and on the probability of given deviations from these normal values. These data are included for the typical Humphrey Field Analyzer fields, but you need to collect this information if you use other perimeters

@@

Note that you have to specify what type of data is contained in the visual field tables. The default is set to *total deviation*.

@@

vfplot(example2, type = "td")

@@

@@

@@

# Global parameters

Global parameters summarize the visual field in a few parameters. These global parameters also can be calculated with the visualField package. The resulting table contains the visual field header data (patient id, eye, etc.), but contains 8 columns (msens, ssens, tmd, tsd, pmd, psd, gh, and vfi) instead.

The mean sensitivity (MS, msens) is the arithmetic mean of all sensitivity values, the mean deviation (MD, tmd) is the arithmetic mean of all total deviation values in the field. The pattern standard deviation (PSD) is the standard deviation of the total deviation values.

@@

getgl(example2)[, -(1:getlocini()-1)]

```
Again, the probability of the observed values given that the visual field is normal can be calculated.

@@

getglp(getgl(example2))[, -(1:getlocini()-1)]

@@

# Structure function correlations in glaucoma patients

In the last years, techniques detailed imaging of the retina and the optic nerve have made incredible advances. Therefore, correlation between structure and function is becoming even more important than ever.

When photoreceptors are lost in an specific retinal area, the scotoma always occurs in the directions in space that is perceived by that retinal area. This makes structure-function correlations easy.

In glaucoma, the damage occurs in the optic nerve at the levels of the nerve fiber bundles, which consist of the axons of the retinal ganglion cells. In the retina, these cells form two layers: the ganglion cell layer that contains the cell bodies, and the retinal nerve fiber layer. However, the cell bodies of the retinal ganglion cells are displaced from the area where the photoreceptors are located that they receive input from. This displacement is relevant mostly in the center of the visual field.

Therefore, the visualFields package provides function to calculate where the ganglion cells that subserve a given location in the visual field are located and to calculate at which angle the axons of these ganglion cells enter the optic nerve head (optic disc), where they pass through the sclera (wall of the eye) to form the optic nerve.

>>>>

---
title: visualFields 1.x
author:
  - Cord Huchzermeyer, INSTITUTION
  - Iván Marín-Franch, Computational Optometry
date: "`r Sys.Date()`"
output: html_document
bibliography: references.bib
biblio-style: "apalike"
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The Open Perimetry Initiative (OPI)
The OPI [@IMF:Tur2012] [@IMF:Marin2022a] is a standard for interfacing with visual field testing machines (perimeters). It specifies basic functions that allow many visual field tests to be constructed. As of October 2017 it is fully implemented on the Octopus 900 and partially on the Heidelberg Edge Perimeter, the Kowa AP 7000, the CrewT IMO and the Centervue Compass.

The visualFields package, part of the OPI, is a collection of tools for analyzing the field of vision. It provides a framework for the development and use of innovative methods for visualization, statistical analysis, and clinical interpretation of visual-field loss and its change over time.

To install the released version of the package, run 
```{r eval=FALSE}
install.packages("visualFields")
```

To install the newestversion of the package, run:
```{r eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("imarinfr/vf1/source")
```

To load the package run
```{r}
library(visualFields)
```

```{r echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(plotrix)
```

# What is the visual field?
The visual field of an eye is the extent of the world that can be seen. It refers to both the central and the peripheral vision. It is measured in degrees of visual angle, with the center of vision being $0^\circ$, points on the temporal and superior regions of the visual field being positive angles and those in the inferior and nasal being negative. On healthy eyes, the sensitivity to light contrast changes as the magnitude of the visual angle increases (or as it is commonly said in perimetry, eccentricity). The difference in light sensitivity over the visual field as a function of eccentricity is known as the Traquair's *hill of vision* [@IMF:Traquair1938]. The figure below shows the hill of vision as rendered by Aulhorn & Harms [@IMF:Aul1967]. Notice in the discontinuity in the hill of vision in the lower right panel. Sensitivity in this region, centered at approximately $(15^\circ, -3^\circ)$ in the visual field is is completely blind, as it corresponds to the anatomical region of the retina known as the *blind spot*, where the optical fiber exits from the eye towards the visual cortex.

<figure>
  <center><img src="./AulhornHillOfVision.png" alt="Hill of Vision" width="75%"></center>
  <center><figcaption>Aulhorn and Harms' rendering of Traquair's hill of vision.</figcaption></center>
</figure>

# What is visual field testing?
Visual field testing measures a patient's visual sensitivity at different locations of the visual field. Its purpose is to reconstruct as best as possible within the confines and limitations of a clinical setting the *hill of vision*. The most common type of visual field testing is *static automated perimetry (SAP)*, which is performed with in a dome-shaped device, so that the distance between the stimulus and the eye equal for all locations within the visual field. [I WOULD DELETE THE FOLLOWING AS IT MAY CONFUSE RATHER THAN CLARIFY GIVEN THAT NOBODY USES CAMPIMETRY ANYMORE, AT LEAST IN THE UK OR US] In contrast, *campimetry* is performed with a flat screen.

Usually, perimetry tests the ability to detect increments in stimulus luminance with respect to the background with an achromatic (not colored) stimulus. This is a form of perimetry known as *white-on-white perimetry*. However, theoretically, other aspects of vision can be tested at different locations (ability to discriminate dots that are very close to each other, ability to discriminate similar colors, contrast sensitivity). The figure below shows the most common grid of locations tested during *SAP*. The gray ellipse corresponds to the blind spot of the hill of vision and, therefore, the red dots are expected to have no sensitivity.

@@

plot(x = default_locmap$coord$x, 
     y = default_locmap$coord$y, 
     type = "n", 
     asp = 1,
     xlim = c(-30, 30),
     ylim = c(-34, 34),
     yaxt = 'n',
     xaxt = 'n',
     bty = 'n',
     xlab = "",
     ylab = "",
     main = "24-2 grid of test locations")
lines(-c(30, 2.5), c(0, 0), lwd = 2, col = "lightgray")
lines(c(2.5, 30), c(0, 0), lwd = 2, col = "lightgray")
lines(c(0, 0), -c(30, 2.5), lwd = 2, col = "lightgray")
lines(c(0, 0), c(2.5, 30), lwd = 2, col = "lightgray")
lines(x = c(-2.5, 2.5), y = c(0, 0), lwd = 1.5)
lines(x = c(0, 0), y = c(-2.5, 2.5), lwd = 1.5)
for(r in c(10, 20, 30))
  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
draw.ellipse(15, -1.5, 2.75, 3.75, col = "gray", border = NA)
text(x = 0, y = -34, "horizontal eccentricity [degrees]")
text(x = -34, y = -0, "vertical eccentricity [degrees]", srt = 90)
text(x = 31, y = 0, "30\u00b0", adj = c(0, 0.5))
text(x = 0, y = 31, "30\u00b0", adj = c(0.5, 0))
points(x = default_locmap$coord$x[-default_locmap$bs],
       y = default_locmap$coord$y[-default_locmap$bs])
points(x = default_locmap$coord$x[default_locmap$bs],
       y = default_locmap$coord$y[default_locmap$bs],
       col = "red", pch = 19)

@@

The visual fields package was created to analyze data from *SAP*. In SAP, the goal is to quantify sensitivity by increasing and decreasing the light differential (or contrast) between stimulus and background and asking the observer to press the button when they detect the presence of the visual stimulus. The contrast threshold is determined at each location of the visual field as the one for which the observer is able to detect the presence of the stimulus 50% of the times. Such thresholds are estimated at a fixed number of locations forming a regular grid as the one in the figure above. There is another common form of perimetry, *kinetic perimetry*, where a light spot is moved from the periphery towards the center along a number of meridians. In this form of perimetry the location where the spot is first detected by the subject is noted and those points at different meridians are connected to find isopters of equal sensitivity on the retina. The statistical analysis for this form of perimetry is very different than for static perimetry and beyond the scope of this visualFields package.

# Performing visual field tests
For the typical visual field test, an observer (usually a patient in an ophthalmology or optimetry clinic) is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld dome (or hemisphere). The perimeter is located in a dimly lit, calm room and the observer has been explained what to do by an experienced technician before the examination. In his hand, the observer has access to a clicker.

Observers now look at a fixation target (a dot, a cross, a diamond, or other) right in front of them in the center of the dome and are presented visual stimuli over their visual field. Observer are to press the clicker any time a stimulus appears anywhere in the dome These lights are presented only for a short time span (typically 100-200ms), so that the observer has no time to shift his gaze to the stimulus. A press of the clicker will be linked to a presentation when it occurs within a given time window after the presentation. A computer will present stimuli several times at different locations in a grid pattern and in a spatially random order. For each location, the stimulus contrast will be decreased if the observer responded to the previous response and increased if not.

At the end of the test, the inverse of the contrast threshold, $\Delta L_{\rm threshold}$ at each location is recorded in a decibel scale (dB). This dB scale is normally defined as attenuation from the maximum stimulus possible $\Delta L_{\rm max}$ as:
\begin{equation}
dB = -10\ \rm log \left (\frac{\Delta L_{\rm thr}}{\Delta L_{\rm max}} \right)\ .
\end{equation}
For the most common devices, The Humprhery Field Analyzer and the Octopus 900, the maximum luminance differential $\Delta L_{\rm max}$ is $10000\ (\rm apostilb) / \pi = 3183.1\ cd\ m^{-2}$.

These are the data expected by the visualFields package for statistical analysis and generation of visualization devices.

## Grid patterns
The spatial position of the points $l_1$ to $l_k$ [CAN WE USE K FOR LOCATIONS, TO PROTECT N FOR NUMBER OF VISUAL FIELDS OR SAMPLE SIZE?] in the visual field is found in the location map (locmap) and depends on the pattern used by the perimeter. When analyzing data, you need to specify the locmap.

By default, the 24-2 pattern (shown in the previous figure) is used, which consists of 54 locations that are equally spaced across the field.

```{r}
str(default_locmap)
```

Beside a name and a description, a locmap dataset contains a coordinate table with the x- and y-coordinates of the locations measured in degrees, and the id of the points that cover the physiological blind spot (the id is the number x in $l_x$ and corresponds to the row in the coord table).

Other grid patterns are also defined in the package. For instance, the 10-2 grid:

```{r}
ten_two_locmap <- locmaps$p10d2
str(ten_two_locmap)
```

And many others:

```{r}
names(locmaps)
```

The grid, the stimulus size, presentation time and the algorithm responsible for varying intensities remain constant throughout the examination and are called a perimetry program. Not that examinations performed with different programs cannot be directly compared.

# visualFields as an aid for clinical assessment of the visual field
In clinical practice, visual field testing is used identify decreases in visual function caused by disease. Such *visual field defects* or *scotomas* can be caused by diseases that affect tissue anywhere along the optic pathway. This ranges from the optic media (cornea / lens / vitreous) which transmit light, to the retina, which detects lights and preprocesses visual information, to the optic nerve, which transmits information, and up to the visual cortex.

Often, clinicians can draw reliable conclusions about the localisation of the disease by the pattern of the visual field defects. For example, an infarction of the right visual cortex leads complete blindness in the left hemispheres of both eyes with a very distinct border along the vertical meridian. Therefore, the patient does not perceive anything to the right of the point they look at. On the other side of the vertical meridian the visual field is normal. In contrast, patients with glaucoma start with areas of slightly decreased light sensitivity often in the upper or lower hemifield with a bow-shaped (arcuate) pattern cause by the localisation of damage in the optic nerve head.

Visual field tests can be used to diagnose disease, to appraise the functional consequences of disease (for example: can the patient still drive a vehicle?), and to monitor progression.

[THIS SEEMS TO BE A BIT OUF OF PLACE, BUT I PREFER NOT TO EDIT THIS PART MYSELF]
The most advanced statistical methods have been created for monitoring glaucoma progression. The open-angle glaucomas are a group of slowly progressing, chronic neurodegenerative diseases of the optic nerve, often caused by increased eye pressure. Patients need consequent monitoring over many years and slight changes in the visual field must be detected reliably. Therefore, many of the functions of the visualField package focus on glaucoma research.

# Getting started with the package
## Loading data
The visualField package contains some example data. This dataset contains the results of 42 visual field tests of one patient (id: sample1). The patient has primary open-angle glaucoma (type: pwg).

@@

@@

Each row corresponds to one visual field test. Visual field testing is performed monocularly (one eye at a time), so that two examinations exist for each date. The abbreviations for the laterality ('eye') is derived from latin (OD: right eye = oculus dexter, and OS: left eye = oculus sinister), when both eyes are examined, the abbreviation is OU.

@@

vfpwgSunyiu24d2[1, getvfcols()]

@@

## Plotting the visual field
The most straightforward plots show the sensitivities at the corresponding locations. These representations are very exact, but may be difficult to interpret. For learning purposes, here, we manually plot some visual field data.

@@

setdefaults()
###[I HAVE CHANGED THE EXAMPLE TO ONE FROM THE RETEST DATASET BECAUSE THE ONE FROM SUNY/IU IS TOO EXTREME]

example2 <- vffilter(vfpwgRetest24d2, 
                     id == "1" & 
                       eye == "OD")[1,]

@@

example2_dataframe <- data.frame(x = getlocmap()$coord$x, 
                                 y = getlocmap()$coord$y, 
                                 td = as.vector(unlist(example2[,getvfcols()])))

ggplot(example2_dataframe, aes(x = x, y = y)) +
  geom_text(aes(label = td)) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  coord_fixed() +
  theme_void()

Note that left eyes have to be treated differently if you plot the points manually, because the location labels are mirrored along the y axis. In other words, to plot the data manually, for a left eye x should be `-getlocmap()$coord$x`.

@@

ggplot(ODOS_dataframe, aes(x = x, y = y)) +
  coord_fixed() +
  xlim(-30, 30) +
  ylim(-24, 24) +
  geom_label(aes(label = loc_label, color = is_blindspot), size = 2.75) +
  scale_color_manual(values = c("black", "red")) +
  facet_wrap(~ eye)

@@

# Psychophysical testing
[THIS SECTION COULD BE MERGED WITH "Performing visual field tests"]
Psychophysical tests systematically measure the relationship between physical stimulus strength and the subjective perception. This is performed by varying the stimulus and asking the observer to communicate his perception with a limited number of allowed responses. In visual field testing, the observer usually presses a button when a light spot is perceived.

@@

# Measurements
[THIS SECTION COULD BE AFTER "Performing visual field tests"]
Although usually not seen in visual field data, the most elementary unit of a static visual field test is a subject's response. It consists of the constant stimulus parameters (size, presentation time, background luminance), the stimulus location, the stimulus intensity and whether the subject pressed the button within a predefined time window to indicate that he saw the stimulus.

@@

## False negatives and false positives
[THIS SECTION COULD BE AFTER "Performing visual field tests" TOO?]

@@

# Mathematical derivatives of contrast thresholds for statisitcal analysis
## Identifying defective locations on the visual field
[A PART OF THIS HAS BEEN EXPLAINED IN MY EDITED VERSION OF "Performing visual field tests". I HAVE REWORKED THIS SECTION. SEE IF MY EDITS MAKE SENSE]
As mentioned earlier, the sensitivity in dB of attenuation from the maximum stimulus at each location is computed from the contrast thresholds.

The sensitivity to light fluctuates considerably between measurements in the same observer (test-retest variability ---and more as the disease progresses) and between observers (inter-individual differences). They also decrease at different rates depending on the eccentricity with age.

With these predictable and unpredictable differences in the pointwise sensitivity measurements, how can we provide tools and reports to aid the clinical assessment of the visual field? Or more to the point, how can we discriminate between healthy locations and locations that are damaged or suspected of being damaged? Clinically it is more useful to compute and report the deviations from an age-corrected mean normal visual field, since it becomes easier the identification of spatial patterns of damage thus facilitating the clinical assessment of the visual field.

But this analysis requires the use of a *normative dataset* from which to obtain models describing the age effect on sensitivity at each location. These functions are usually linear and those *age linear models* are used to obtain the *age-corrected mean normal visual field*.

The resulting *deviations* from age-corrected mean normal visual field are also measured in dB and they are interpreted as follows. In the Humphrey Field Analyzer, a negative deviation value means that the sensitivity at that location is below mean normal. By contrast, on Octopus perimeters, negative is above and positive below mean normal. In this package, those differences are called *total deviation* and their interpretation are as for the Humphrey perimeter. In case you use visualFields with data from a different perimeter it is important to know what is the definition of *deviation* and change the sign if necessary.

In eye diseases like glaucoma, general defects that affect the whole visual field are often less relevant than focal scotomas. And often ocular opacities caused by cataract can mask such focal defects. For this reason, a statistical post-hoc correction to *total deviation* values is further applied to obtain *pattern deviation* values. The correction factor is often called the *general height* and for the 24-2 Humphrey is obtained by selecting the 7th greatest total deviation value. The *pattern deviation* values equal the *total deviation* values minus the *general height*.

## Normative dataset and normative references
The visualFields package contains three datasets of healthy subjects that can be used to obtain normative age linear models and references.

The SUNY-IU dataset [@IMF:Mar2013b] for the 24-2 SITA Standard and the Goldmann Size III:
```{r SUNYIU datasets}
str(vfctrSunyiu24d2[,1:10])
```

The Iowa dataset [@IMF:Marin2018c] for a bimodal ZEST algorithm and the Goldmann Size V:
```{r Iowa datasets}
str(vfctrIowaPC26[,1:10])
str(vfctrIowaPeri[,1:10])
```

From these datasets, or any other dataset of healthy eyes, normative references can be computed as follows:

[CORD: I ADDED THIS TO SHOW THAT VISUALFIELDS IS NOT JUST TO ANALYZE THE DATA IN A CERTAIN WAY, YOU GENERATE YOUR OWN NORMATIVE VALUES AND WORK WITH THEM. SO FROM HERE ALL EXAMPLES ARE FROM nv INSTEAD OF getnv(). AT THE END OF THE DOCUMENT, WE CAN EXPLAIN setnv(), getnv(), setlocini(), setdefaults(), ETC. OK?]

```{r nv}
nv <- nvgenerate(vfctrSunyiu24d2)
```

For future use, we set `nv` as the normative values to be used in visualFields for the calculation of total and pattern deviation values and other statistical analyses.
```{r setnv}
setnv(nv)
```

### Age linear models and mean normal visual fields
The normative values and references stored in `nv` contains the coefficients of the age linear models:

```{r agelm}
print(nv$agem$coeff)
```

And a function from where the age-corrected mean normal visual fields can be computed:
```{r agemeannormal}
age <- 50
print(nv$agem$model(age))
```

Notice that the locations corresponding to the anatomical region of the blind spot all have `NA` as values, as they are excluded from the statistical analysis.

## Total deviation
Total deviation is calculated from the sensitivities and from age-related normal values. These are saved in the environment variable *nv* (normal values).

```{r sensitivities}

print(example2[, getvfcols()[1:10]])

@@

@@

age <- example2$age[1]

@@

normal_values <- getnv()$agem$model(age)[1:10]

@@

total_defects <- example2[, getvfcols()[1:10]] - normal_values

@@

td <- gettd(example2)
print(td[, getvfcols()[1:10]])

@@

## Pattern deviation
Pattern deviation can be calculated after obtaining the general height.
```{r gh}
gh <- getgh(td)
print(gh)
pattern_defects <- td[, getvfcols()[1:10]] - gh
print(pattern_defects)
```

@@

pd <- getpd(td)
print(pd[, getvfcols()[1:10]])

@@

## Total and Pattern deviation probabilities
As mentioned psychophysical sensitivity values vary considerably from test to test and between subjects Therefore, even in perfectly average eyes deviations will be different from zero most of the time. Given a visual field from a patient, the clinician has to know if a given deviation is likely to be caused by disease as opposed to normal retest and inter-individual variability.

The function `nvgenerate` computes for the normative dataset, the total deviation and pattern deviation values for healthy eyes and records pointwise empirical probability distribution functions [@IMF:Hei1987b] that can be used to compute the probability that a specific total deviation (or pattern deviation value) value is from a healthy location [@IMF:Hei1987a] [@IMF:Heijl1991]. Thus, a low probability implies a greater likelihood of the location being damaged. Typically (and in our example), the probabilities are categorized as
```{r probability categories}
print(getnv()$lut$probs)
```

where `0.005` means probability $< 0.005$, where `0.01` means probability $< 0.01$, etc. A probability greater than 0.05 is considered within the normal range.

The probability values for the previous values are as follows.

```{r probability maps}
print(gettdp(td)[, getvfcols()[1:10]])
print(getpdp(pd)[, getvfcols()[1:10]])
```

# Plotting the visual field
There are different ways to visualize the perimetric results. These visualizations have to be interpreted like maps in that they represent a simplified version of the hill of vision.

[WE PROBABLY NEED SOME INTRO TO VFPLOT BEFORE TALKING MAKING NOTES, ETC?]
The standard already contains some additional information with a border presenting a color-coded visualization of the deviation and a graphical representation of the blind spot. Note that the vfplot with the option "td" needs information on differential light-sensitivities in a normal reference population and on the probability of given deviations from these normal values. These data are included for the typical Humphrey Field Analyzer fields, but you need to collect this information if you use other perimeters

Note that you have to specify what type of data is contained in the visual field tables. The default is set to *total deviation*, `td`.

vfplot(example2)

@@

In this graph, yellow represents probabilities < 0.05, orange probabilities < 0.02, red < 0.01, and dark red < 0.005.

It is possible to print sensitivities (with darker representing locations with lower sensitivities)

@@

Pattern deviation values

@@

@@

And an hybrid plot showing total deviation values and probabilities along with a grayscale representing the sensitivities.

@@

And a similar hybrid plotbut with pattern deviation values and probabilities.

@@

# Global indices
Global indices summarize the visual field in a few parameters and their values and probability level can be calculated with the visualField package. The resulting table contains the visual field header data (patient id, eye, etc.) and 8 additional columns (msens, ssens, tmd, tsd, pmd, psd, gh, and vfi).

@@

print(getgl(example2)[,-(1:getlocini()-1)])

The column `msens` is the mean sensitivity (MS); the arithmetic mean of all sensitivity values. The column `ssens` is the standard deviation of sensitivity values.

The column `tmd` is the mean deviation (MD); the arithmetic mean of all total deviation values. The column `tsd` is the standard deviation of total deviation values.

The column `pmd` is the mean of pattern deviation values. The column `psd` is the standard pattern deviation (PSD) or the standard deviation of the pattern deviation values.

The last column, `vfi`, is the visual field index (VFI) [@IMF:Ben2008], and the second to last is the general height (GH).

The probability of the observed values given that the visual field is normal can also be calculated.

@@

print(getglp(getgl(example2))[, -(1:getlocini()-1)])

@@

# Putting everything together.
Single field analysis reports can be created as pdfs or as interactive reports with Shiny. To explore them, type in the R Console the following:

```{r eval = FALSE, include = TRUE}
vfsfa(example2, file = "test.pdf")
vfsfashiny(example2)
```

# Structure function correlations in glaucoma patients
In the last years, techniques that provide detailed imaging of the retina and the optic nerve have made incredible advances. Therefore, correlation between structure and function is becoming even more important than ever.

When [SINCE THIS IS MORE FOCUSED TO GLAUCOMA, I REMOVED PHOTORECEPTORS AND PUT INSTEAD GANGLION CELLS, AS PHOTORECEPTORS REMAIN INTACT IN GLAUCOMA] ganglion cells are lost in characteristic patterns at the retinal, [I DO NOT UNDERSTAND THIS SENTENCE] the scotoma always occurs in the directions in space that is perceived by that retinal area.

To aid comparisons between structural insult and functional defect, it is useful to understand and model the anatomy of a typical healthy eye, even if such modeling through is imperfect given the large inter-individual differences in anatomical features [@IMF:Jan2009] [@IMF:Jan2012] [@IMF:Drasdo2007] [@IMF:Montesano2020]

In glaucoma, the damage occurs in the optic nerve at the levels of the nerve fiber bundles, which consist of the axons of the retinal ganglion cells. In the retina, these cells form two layers: the ganglion cell layer that contains the cell bodies, and the retinal nerve fiber layer. However, the cell bodies of the retinal ganglion cells are displaced from the area where the photoreceptors are located that they receive input from. This displacement is relevant mostly in the center of the visual field, so it would greatly affect the structure-function comparisons on grids such as the 10-2. The following figure shows the visual field locations and the corresponding ganglion cell soma being stimulated.
